---
title: "CIW Tutorial 1"
author: "Tina Rozsos"
institute: "VU Amsterdam"
date: "2025-10-31"
date-format: long
format:
  revealjs:
    theme: default
    logo: assets/VU_logo.png
    footer: "CIW Math refresher"
    transition: fade
    controls: true
    embed-resources: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

```{r}
library(tidyverse)
library(patchwork)
theme_set(theme_minimal(base_size = 20))
```

## Introduction

### Overall goal

- Who I am
- Purpose of these tutorials
- Focus is on
  - intuitive understanding of concepts
  - ability to work with economic models, think about practical applications

::: {.fragment}
### Plan for today

1. Solve and discuss exercises from first assignment
1. Work on questions from extra materials
1. Open floor
:::

## Assignment 1

Discuss slides XXX

## Attendance

On Canvas go to Assignments > Week 1 - Attendance

Get attendance grade by answering with the number written on the board

## Extra Lecture materials

### Work in groups of 2-4
### On screen 2, choose option 1

## CIW Extra Lecture 1 Screen 6

::: {.incremental}
1. Look at the utility function: why do you think it's called quasilinear?
1. What can we say about the consumer's preferences for goods $x$ and $y$ when $p_x=p_y$?
1. What if income changes but relative prices stay the same?
1. What if relative prices change? (solve the assignment on screen)
1. Would you reach a similar conclusion for the income and substitution effects for $y$ if the price of $y$ changed?
:::

## CIW Extra Lecture 1 Screen 8

::: {.incremental}
1. What is the relationship between gas and food on the graph?
1. What happens if a person is given more gas but the same amount of food?
1. Is this relationship realistic?
1. What is the compensating variation for a change in gas price? (solve the assignment on screen)
1. Would the CV increase, decrease, or stay the same if the goods were $imperfect$ complements?
:::

## Open floor