---
title: "CIW Tutorial 2"
author: "Tina Rozsos"
institute: "VU Amsterdam"
date: "2025-11-07"
date-format: long
format:
  revealjs:
    theme: default
    logo: assets/VU_logo.png
    footer: "Choices, Inequality, and Welfare"
    transition: fade
    controls: true
    embed-resources: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

```{r}
# library(tidyverse)
# library(patchwork)
# theme_set(theme_minimal(base_size = 20))
```

## Plan for today

1. Discuss exercises from the second assignment
1. Data analysis with Jamovi
1. Work on questions from extra materials
1. Open floor


## Assignment 2

### Reminder

- Desmos assignments are partially auto-graded, but not perfectly
    - If Desmos says you did not pass but Canvas says you did: **you pass**
    - If Desmos says you passed but Canvas says you did not: **should not happen, email me**

### Discuss slides 11--12

## Attendance

On Canvas go to Assignments > Week 2 - Attendance

Get attendance grade by answering with the number written on the board

## Data analysis with Jamovi

::: {.incremental}
1. Download panamadata.xlsx from Canvas (Week 2 > Panama Consumption Data)
1. Open Jamovi and read in the data
1. Explore the data:
    1. visualize the distribution of per-capita consumption (*cons_pc*)
    2. visualize the distribution of log per-capita consumption (define new variable *lncons_pc*)
1. What is the headcount ratio (fgt0)?
    1. make new variable *belowz*: is person below povertyline?
    1. get summary statistic of new variable: mean of *belowz* is HC
:::

## Extra Lecture 3 Screen 6-7

### Comparing tax/transfer systems

::: {.incremental}
- What does a worker want from a tax/transfer system?
- What does the government want?
- Any other aspects you that should be considered when evaluating a tax/transfer system?
:::

::: {.fragment}
What would be reasons for/against the following systems?
:::

::: {.incremental}
1. Fixed hourly wage, no taxes or transfers
1. <span style="color: red;">Flat income tax: 16.67% tax on total earnings</span>
1. <span style="color: blue;">Universal basic income, funded by 50% flat income tax</span>
1. <span style="color: green;">Earned-income tax credit based on hours worked</span>
:::

::: {.fragment}
*If time permits:* start sketching out budget lines and optimal choices per system on Amplify
:::

## Open floor - any questions?