---
title: "CIW Tutorial 3"
author: "Tina Rozsos"
institute: "VU Amsterdam"
date: "2025-11-14"
date-format: long
format:
  revealjs:
    theme: default
    logo: assets/VU_logo.png
    footer: "Choices, Inequality, and Welfare"
    transition: fade
    controls: true
    embed-resources: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

```{r}
# library(tidyverse)
# library(patchwork)
# theme_set(theme_minimal(base_size = 20))
```

## Plan for today

1. Discuss exercises from the second assignment
1. Work on questions from the mock midterm
1. Open floor


## Assignment 3

### Discuss slides 6--9, 11

## Attendance

On Canvas go to Assignments > Week 3 - Attendance

Get attendance grade by answering with the number written on the board

## Mock midterm

Which question would you like to discuss?

1. Labor supply: 
  - budget constraints with change in wage rate
  - decomposition of income and substitution effects
2. Inequality, Poverty, and Development:
  - calculating FGT poverty measures
  - Lorenz curve and gini coefficient
3. Public goods:
  - Drawing production possibility frontier and indifference curves
  - Samuelson rule

## Open floor - any questions?