---
title: "CIW Tutorial 4"
author: "Tina Rozsos"
institute: "VU Amsterdam"
date: "2025-11-21"
date-format: long
format:
  revealjs:
    theme: default
    logo: assets/VU_logo.png
    footer: "Choices, Inequality, and Welfare"
    transition: fade
    controls: true
    embed-resources: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

```{r}
# library(tidyverse)
# library(patchwork)
# theme_set(theme_minimal(base_size = 20))
```

## Plan for today

1. Reflection on the midterm
1. Discuss exercises from the assignment
1. Work on questions from the extra materials
1. Open floor

## Reflection on the midterm

::: {.incremental}
- How did it go?
- Any questions/question types you would like to discuss?
- How can the tutorials best help you prepare for the final exam?
:::

## Assignment 4

### Discuss slides XXX

## Attendance

On Canvas go to Assignments > Week 4 - Attendance

Get attendance grade by answering with the number written on the board

## Extra materials 7

### Best responses in Cournot duopoly

### Screens 5-6

Focus exclusively on the perspective of Firm B

::: {.incremental}
- If B could choose A's quantity, what would it choose?
- If B knows that $Q_A=0$, how will B behave in response?
- If A produces a large quantity, how will B respond?
- Is B indifferent between the points on its best response curve? (answer the question on the left panel of screen 6)
- Is a point on the best response curve better than a point outside it? (answer the question on the right panel of screen 6)
:::

## Open floor - any questions?