---
title: "Measuring inequality with the Gini coefficient"
subtitle: "SBE PhD microteaching session"
author: "Tina Rozsos"
institute: "VU Amsterdam"
date: "2024-11-04"
date-format: long
format:
  revealjs: 
    theme: default
    logo: assets/VU_logo.png
    transition: fade
    controls: true
    embed-resources: false
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, error = FALSE,
                      fig.height = 6, fig.width = 6, fig.align = "center")
```
```{r}
library(tidyverse)
theme_set(theme_minimal(base_size = 16))
```

##

::: {.r-fit-text}
Please open the workshop handout at
:::
::: {.r-fit-text}
[tinarozsos.github.io/CIW2024/handout](https://tinarozsos.github.io/CIW2024/handout)
:::

# Income inequality around the world

## 

<iframe src="https://ourworldindata.org/grapher/income-share-distribution-before-tax-wid?tab=chart" loading="lazy" style="width: 100%; height: 600px; border: 0px none;" allow="web-share; clipboard-write"></iframe>

# A measure of inequality: the Gini coefficient

## What is the Gini coefficient?

- Measures inequality in any distribution
- $G=0$ means perfect equality (everyone has the same income)
- $G=1$ means perfect inequality (one person has all the income)

## How to calculate the Gini coefficient?

1. Sort incomes from lowest to highest
2. Plot the cumulative share of income against the cumulative share of the population (Lorenz curve)

::: {.fragment}
Example income distribution in a country with 3 people:

```{r}
d <- tibble(person = c("X", "Y", "Z"),
       income = c(100, 600, 300),
       rank = c(1, 3, 2))
knitr::kable(d)
```
:::

## How to calculate the Gini coefficient?

::: {.columns}
::: {.column width="50%"}
Cumulative income shares:

```{r}
d |> 
  arrange(rank) |> 
  mutate(`cumulative<br>income` = cumsum(income)) |> 
knitr::kable()
```
:::

::: {.column width="50%"}
::: {.fragment}
```{r}
labs <- tibble(
  x = c(0.5, 0.4, 0.7),
  y = c(0.54, 0.3, 0.25),
  label = c("Line of perfect equality", "A", "B"),
  angle = c(45, 0, 0)
)

p <- bind_rows(
  tibble(rank = 0, income = 0), d) |> 
  arrange(rank) |> 
  mutate(cum_inc = cumsum(income) / sum(income),
         cum_pop = rank / max(rank)) |> 
  ggplot(aes(x = cum_pop, y = cum_inc)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_line() +
  geom_point(size = 2) +
  geom_point(aes(y = cum_pop), size = 2, shape = 21, fill = "white") +
  geom_text(data = labs[1,], aes(x = x, y = y, label = label), angle = labs$angle[1]) +
  labs(x = "Cumulative share of population", y = "Cumulative share of income") +
  coord_cartesian(expand = FALSE) +
  scale_x_continuous(breaks = c(0, .33, .66, 1), labels = scales::percent,
                     minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0, .1, .33, .4, .66, 1), labels = scales::percent,
                     minor_breaks = NULL) +
  theme(aspect.ratio = 1)
p
```
:::
:::
:::

## How to calculate the Gini coefficient?
::: {.columns}
::: {.column width="50%"}
Cumulative income shares:

```{r}
d |> 
  arrange(rank) |> 
  mutate(`cumulative<br>income` = cumsum(income)) |> 
knitr::kable()
```

$$G = \frac{A}{A+B}$$
:::
::: {.column width="50%"}
```{r}
p +
  geom_ribbon(aes(ymin = cum_inc, ymax = cum_pop), fill = "steelblue", alpha = 0.75) +
  geom_area(fill = "grey", alpha = 0.75) +
  geom_text(data = labs, aes(x = x, y = y, label = label), angle = labs$angle)
  
```
:::
:::

## Try it for yourself

Use [this](https://www.econgraphs.org/projects/core/unit1/gini.html) Gini coefficient calculator for a sample population with

- $n=1000$ people (adjust slider on right)
- 800 people earn 50K (adjust the bars on the left figure)
- 200 people earn 200K (adjust the bars on the left figure)

::: {.incremental}
1. What is the Gini coefficient for this population?
2. If the 200 people earn 100K instead of 200K, will the Gini coefficient increase or decrease?
3. What if instead 600 people earn 50K and 400 people earn 200K? Why?
:::

# Income inequality around the world (again)

## 

<iframe src="https://ourworldindata.org/grapher/economic-inequality-gini-index?time=latest&tab=map" loading="lazy" style="width: 100%; height: 600px; border: 0px none;" allow="web-share; clipboard-write"></iframe>

## Take a minute to reflect

::: {.absolute top=200 style="font-size: 40px;"}
- Do you find the Gini coefficient a good tool for measuring income inequality?
:::
::: {.absolute top=400 style="font-size: 40px;"}
- Do you see any limitations to the Gini coefficient?
:::

## 

::: {.r-fit-text .center .absolute top=200}
Thank you!
:::